Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
Minor fixes
in a mess

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\quand}{\quad\mbox{and}\quad}

\newcommand{\qquand}{\qquad\mbox{and}\qquad}

@@

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a
  derivation $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$.
  A subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  been introduced by the same instance of $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now we show that
  $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  delete all atoms in the scope of the corresponding quantifier,
  because $\phi'$ is a fibration on the binding graph. Hence, each line
  in $\PE\Deri_1$ is the propositional encoding $\PE P$ of a
  first-order formula $P$.  We now have to show
  that each instance of $\wrD$ is indeed a correct
  application in first-order logic. For this we have to inspect the
  cases a weakening happens inside a subformula $x\vlor C$ or $x\vlan
  C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur.
  Thus we have a first order derivation
  $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$ with $\PE B_1=\PEp B_1$.
\end{proof}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\quand}{\quad\mbox{and}\quad}
\newcommand{\qquand}{\qquad\mbox{and}\qquad}

\newcommand{\quor}{\quad\mbox{or}\quad}
\newcommand{\qquor}{\qquad\mbox{or}\qquad}

@@

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
   $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$. In order to
   lift $\DDeri$, we might need to reorganize the instancese of
   $\wD$. If $H$ contains a subformula $\forall x.A$ which is not
   present in $G$, the $\wD$-instances in $\DDeri$ could introduce the
   parts of the propositional encoding $x\vlor A$ independently.  We
   say that an instance $\rr_1$ of $\wD$ in $\Deri$ is \emph{in the
   scope} of an instance $\rr_2$ of $\wD$ if $\rr_1$ introduced
   formulas that contain a free variable $x$ (i.e., $x$ occurs in a
   term in a predicate) and $\rr_2$ introduces the atom $x$ as a
   subformula (i.e. the propositional encoding of the binder $x$).  We
   can now permute the $\wD$-instances in $\DDeri$ such that whenever
   a rule instance $\rr_1$ depends on an instance $\rr_2$, then
   $\rr_2$ occurs below $\rr_1$ in $\DDeri$.
   We can now lift $\DDeri$ stepwise. First, observe that each line of
   in $\DDeri$ is $\PE\fequ$-equivalent to the propositional encoding $\PE P$ of a
   first-order formula $P$.  We now have to show
   that each other instance of $\wrD$ in $\DDeri$ is indeed the image of a correct
   application of $\wD$ or $\wfaD$ in first-order logic.
   If we have a $\wD$
   \begin{equation*}
     \vlinf{\wD}{}{\PE S\cons{x\vlor \PE A}}{\PE S\cons{\PE A}}
     \qquor
     \vlinf{\wD}{}{\PE S\cons{(x\vlor \PE B)\vlor \PE A}}{\PE S\cons{\PE A}}
   \end{equation*}
   then $x$ cannot occur freely in $A$, as otherwise the fibration
   property would be violated. We can therefore lift these instances to
   \begin{equation*}
     \vlinf{\wfaD}{}{S\cons{\forall x. A}}{S\cons{A}}
     \qquor
     \vlinf{\wD}{}{S\cons{(\forall x. B)\vlor A}}{S\cons{A}}
   \end{equation*}
   respectively.     
  %% A subformula of $\PE B$ is called \emph{weak} if it has been
  %% introduced by a weakening. All subformulas of a weak formula are
  %% also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  %% \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  %% $\Sconhole$. We can assume without loss of generality that
  %% whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  %% been introduced by the same instance of $\wrD$ in
  %% $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  %% brought into this form by simple rule permutations, and then
  %% collapsing several weakenings into a single one.}  Now we show that
  %% $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  %% weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  %% delete all atoms in the scope of the corresponding quantifier,
  %% because $\phi'$ is a fibration on the binding graph. Hence, each line
  If a  weakening happens inside a subformula $x\vlor \PE C$ or $x\vlan
  \PE C$ in $\DDeri$, then there are the following cases:
  \begin{equation*}%\small
    \scalebox{.9}{$
    \vlinf{\wD}{}{\PE S\cons{x\vlor\PE  D\vlor \PE C}}{\PE S\cons{x\vlor\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{x\vlan (\PE D\vlor \PE C)}}{\PE S\cons{x\vlan\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{(x\vlor\PE  D)\vlan \PE C}}{\PE S\cons{x\vlan \PE C}}
    $}
  \end{equation*}
  The first two cases can be lifted to
  \begin{equation*}
    \vlinf{\wD}{}{S\cons{\forall x.( D\vlor C)}}{S\cons{\forall x.  C}}
    \quand
     \vlinf{\wD}{}{S\cons{\exists x.( D\vlor C)}}{S\cons{\exists x.  C}}
  \end{equation*}
  respectively. But in the third case,  an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur. All other situations can be lifted trivially, giving us
  $\vlder{\set{\wrD,\wfaD\fequ}}{\Deri}{H}{G}$ as desired.
\end{proof}
